---
title: "Ametop Alone vs. Ametop and Vapocoolant Spray to Reduce IV Insertion Pain for Pediatric Patients"
subtitle: | 
  |
  | Dr. Louis Scheepers
  | Investigator and Pediatric Anesthesiologist, BC Children's Hospital, Clinical Associate Professor, Department of Anesthesiology, Pharmacology & Therapeutics, Faculty of Medicine, UBC
  |
  | STAT 450: Case Studies in Statistics
  | Instructors: Keegan Korthauer, Melissa Lee, Rodolfo Lourenzutti
  | Teaching Assistant: Jonathan Agyeman
  | Students: Fabiola Grace, Maggie Ruan, Runhe Guo, Yimin You
  | Department of Statistics, The University of British Columbia
  |
date: "15 April 2024"
output: pdf_document
header-includes:
   - \usepackage{float}
     \let\origfigure\figure
     \let\endorigfigure\endfigure
     \renewenvironment{figure}[1][2] {
     \expandafter\origfigure\expandafter[H]
     } {
      \endorigfigure
     }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# Set up echo as false to not print the R code in the report
library(knitr)
library(readxl)
library(ggplot2)
library(broom)
library(tidyverse)
library(gridExtra)
```

\pagenumbering{gobble}

\newpage
\pagenumbering{roman}
\setcounter{page}{2}

# Summary

This study aims to improve pain management for pediatric patients during intravenous (IV) insertions, a common yet distressing procedure. It investigates the effectiveness of two treatments: Ametop gel as a standalone treatment and the combination of Ametop and vapocoolant spray. A randomized controlled experiment is used to assess the pain levels of pediatric patients during IV insertions after receiving either Ametop gel alone or a combination of Ametop gel and vapocoolant spray. A two-sample t-test is used to compare changes in pain levels between the two treatment groups, and a linear regression is performed to examine the association between age and pain level. The two-sample t-test indicates that adding vapocoolant spray significantly reduces the discomfort of IV insertions in pediatric patients. The regression analysis reveals that younger children experience higher levels of pain. Additionally, the use of vapocoolant spray appears to reduce pain more effectively in younger children. The evaluation of the number of IV attempts and side effects shows that adding vapocoolant spray does not increase the number of IV attempts. Overall, the combination of Ametop and vapocoolant spray is more effective than Ametop alone, particularly for younger children.

\newpage
\tableofcontents

\newpage
# List of Illustrations
\listoffigures
\listoftables

\newpage
\pagenumbering{arabic}

# 1. Introduction 

Intravenous (IV) insertion is an essential medical procedure that is frequently associated with discomfort and pain, particularly in pediatric patients. To mitigate this, BC Children’s Hospital currently utilizes Ametop gel, a topical anesthetic that numbs the skin before needle insertion, to alleviate pain and ease the process. Despite its efficacy, Ametop gel alone may not suffice, especially in individuals with heightened sensitivity to pain. The hospital is considering the additional use of a vapocoolant spray. This spray acts rapidly to cool and numb the skin and is hypothesized to provide better pain relief during IV insertions.

To evaluate the effectiveness of adding a vapocoolant spray to Ametop gel, we summarized two research questions and formulated the corresponding statistical analysis. The objectives are as follows:

- Does the addition of a vapocoolant spray to Ametop reduce discomfort in pediatric patients during IV insertions more than Ametop alone? To assess this, the mean difference in pain levels experienced by pediatric patients is examined when using Ametop alone versus the combination of Ametop with a vapocoolant spray during IV insertions.

- Does age and randomization group (control vs. study) affect the pain level during IV insertions? To examine this, the relationship between patient age and randomization group to the pain level is analyzed.

Furthermore, we aim to analyze whether the use of vapocoolant spray affects the number of IV attempts and to assess the incidence of side effects among pediatric patients. The following sections detail the data description, statistical methods, and the results derived from our analysis.

\newpage

# 2. Data Description

Data were collected from 240 pediatric patients undergoing intravenous (IV) insertion. Some participants withdrew from the study for reasons such as itchiness from Ametop and conversion to inhalation induction. We excluded 17 participants who withdrew from the study, focusing solely on the remaining 223 observations.

The dataset has several key variables, categorized as follows:

- **Patient Demographics:** This includes variables such as age and gender. Age (5-16 years old) is particularly important for determining if responses to pain management strategies vary among different pediatric age groups.
- **Randomization Groups:** Patients are randomly assigned to one of two groups: Group 1, receiving Ametop alone (the control group), and Group 2, receiving both Ametop and the vapocoolant spray (the study group). This randomization is essential for assessing the comparative effectiveness of the two treatment approaches.
- **FPS-R Scores:** Patients’ pain levels are rated using the Faces Pain Scales-Revised scale from 0 (no pain) to 10 (worst pain possible), offering a measurable discomfort level during IV insertion. This score determines the primary outcome, that is, the effectiveness of different treatment approaches.
- **Number of IV Attempts:** This variable represents the number of attempts needed for a successful IV insertion, reflecting the procedural difficulty and its potential correlation with patient discomfort.
- **Side Effects:** Side effects, including the appearance of redness, sensations of itchiness, puffiness, or pain, are recorded to check if participants’ experience of any severe adverse reactions to the treatment.
- **Observer Scores:** Observer scores are recorded based on the patients’ observed discomfort, which was evaluated by a nurse using a 3-point scale: 'none,' 'slight pain,' or 'severe pain.' This approach helps validate the primary outcome, the FPS-R scores, by ensuring a reliable assessment of pain in children who may not fully understand the FPS-R pictogram. In our subsequent statistical analysis, we will focus primarily on the FPS-R scores rather than the observer scores.

Other than these variables, there are nine controlled variables, including whether the patient has any Ametop allergies or a needle phobia. These control variables ensure that participants meet the eligibility criteria for the study.

\newpage

# 3. Methods

Several statistical methods are employed to address our research objectives. A two-sample t-test is used to assess the primary objective of comparing the difference in mean FPS-R scores between the control and study groups. For the first secondary objective, a linear regression analysis is performed to explore the relationship between age and randomization groups on FPS-R scores. To illustrate the second secondary objective, a bar graph is presented depicting the number of IV attempts across the two groups. Lastly, a pie chart is used to display the distribution of side effects among all participants, addressing the third secondary objective.

\vspace{3mm}

**3.1 Main Objective: To assess the mean difference in pain levels experienced by pediatric patients when using Ametop alone versus the combination of Ametop with a vapocoolant spray during IV insertions**

The exploratory data analysis examines the distribution of the target response variable, the FPS-R scores. Figure 1 displays comparative boxplots for the control and study groups, illustrating the distribution of these FPS-R scores.

\vspace{3mm}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=1.5, fig.align='center', fig.cap = "FPS-R Scores Boxplots of Control (Ametop) and Study (Ametop + Spray) Group"}
# Read data
file_path <- "../data/Ametop-PE data(new-Alex)(1control 2study).xlsx"
original_data <- read_excel(file_path)

# Save original data for side effects analysis
data <- original_data

# Remove withdrawn patients
data <- data[data$`Withdrawn after consent?` != "Yes", ]

# Remove unnecessary variables
data <- select(data, 
               -`Elective surgery requiring IV?`,
               -`ASA I or II?`,
               -`Aged 5-16?`,
               -`Any allergies to Ametop, Pain Ease, or Tagederm adhesive?`,
               -`Ametop placed at least 30min before estimated IV start time?`,
               -`Receiving sedative pre-medication / anxiolytics?`,
               -`Needle phobia?`,
               -`Planned inhalation induction?`,
               -`Developmental delay or unable to interpret FPS-R?`,
               -`Time Ametop applied`,
               -`Time Ametop removed`,
               -`Time of skin puncture`,
               -`Record ID`,
               -`Complete?`,
               -`Withdrawn after consent?`,
               -`Reason for withdrawal`,
               -`Other reason for withdrawal`)

# Boxplots
plot1 <- data %>% filter(Randomization == 1) %>%
  ggplot(aes(`FPS-R score`, fill = "Control Group")) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Control Group FPS-R Scores Boxplot") +
  scale_fill_manual(values = c("Control Group" = "#219ebc")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 10, face = "bold"), legend.position = "none")

plot2 <- data %>% filter(Randomization == 2) %>%
  ggplot(aes(`FPS-R score`, fill = "Study Group")) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Study Group FPS-R Scores Boxplot") +
  scale_fill_manual(values = c("Study Group" = "#fb8500")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 10, face = "bold"), legend.position = "none")

grid.arrange(plot1, plot2, ncol = 2, nrow = 1)
```

Figure 1 shows that the median FPS-R score for both the control and study groups is 2. The study group displays a higher concentration of scores at FPS-R scores 0 and 2, resulting in a right-skewed distribution, and FPS-R scores 6 and above are considered outliers.

The two-sample t-test is used to assess whether the addition of vapocoolant spray reduces discomfort. This test is applied under the assumption that the differences between the means of the control and study groups follow a normal distribution. Although the dataset itself is not normally distributed, the sample size is large enough that, according to the Central Limit Theorem (CLT), the t-test can still be used. The null hypothesis asserts that the mean FPS-R scores for both groups are the same, while the alternative hypothesis is that the mean FPS-R score is lower for the study group that is using the vapocoolant spray. Consequently, a one-sided t-test is used, and the level of significance is set at 5%.

**3.2 Secondary Objective 1: To determine the relationship between age and randomization (control and study) groups on FPS-R scores**

A linear regression analysis is used to address the second statistical question concerning the impact of age and randomization (study and control) groups on FPS-R scores. This approach explores the relationship between age, the randomization group, and the interaction of both age and the randomization group on the FPS-R scores. To construct the model, the assumptions underlying linear regression are tested. These assumptions include linearity, independence, homoscedasticity, and normality of residuals. The QQ plot shows a heavy-tailed distribution, suggesting the data are not exactly normally distributed. However, given the focus on inference rather than prediction, and based on the Central Limit Theorem, we proceed with using linear regression for our analysis. The statistical significance of the model coefficients is assessed at a 5% significance level.

\vspace{3mm}

**3.3 Secondary Objective 2: To explore the change in the number of IV insertion attempts when using Ametop alone versus the combination of Ametop with a vapocoolant spray**

To investigate whether the combination of Ametop with a vapocoolant spray increases the number of IV insertion attempts compared to the application of Ametop alone, we use summary statistics and visually analyze the data using plots. These help us understand the distribution and average number of IV attempts in both the control group (Ametop alone) and the study group (Ametop with vapocoolant spray).

\vspace{3mm}

**3.4 Secondary Objective 3: To explore the incidence of side effects due to Ametop**

This exploratory data analysis is conducted on all 240 participants of the study to include all patients who experience side effects in the dataset. The observed side effects are then categorized into four distinct groups: “Redness,” “Redness and Itching,” “Puffiness,” and “Pain” to facilitate the quantitative analysis of side effects. The results are summarized with a pie chart and frequency table.

\newpage

# 4. Results

This section presents the results of our data analysis. We compare the p-value to the significance level for the two-sample t-test. Then, we examine a linear regression model on the response variable, using their estimates and p-values. The results are presented in Table 1, Table 2, Figure 2, Figure 3 and Figure 4.

\vspace{3mm}

**4.1 Main Objective: To assess the mean difference in pain levels experienced by pediatric patients when using Ametop alone versus the combination of Ametop with a vapocoolant spray during IV insertions**

For the main objective, a two-sample t-test is performed on the two randomization (control and study) groups and FPS-R scores. The results of the test, including the group means, standard error, and p-value, are shown in Table 1 below.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
table_data <- data.frame(
  `Control Group Mean` = 2.6909,
  `Study Group Mean` = 1.9292,
  `Standard Error` = 0.3779,
  `P-value` = 0.02258,
  check.names = FALSE
)

kable(table_data, caption = "Control Group and Study Group Two-Sample T-Test", format = "markdown")
```

The mean FPS-R score for the group with Ametop alone is 2.6909, while the mean FPS-R score for the group with the combination of Ametop and vapocoolant spray is 1.9292. The test yields a p-value of 0.02258, which is below the 5% significance threshold. This allows us to reject the null hypothesis at the 5% significance level, indicating a difference in the mean FPS-R scores between the control and study groups. These results support the hypothesis that adding vapocoolant spray may reduce discomfort in pediatric patients.

\vspace{3mm}

**4.2 Secondary Objective 1: To determine the relationship between age and randomization (control and study) groups on FPS-R scores**

For the second statistical question, a linear regression model is built using age, randomization group, and the combined effect of both age and randomization group to find their relationship with the FPS-R scores. The results of the linear regression model are shown in Table 2 below.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
regression_table <- data.frame(
  Term = c("(Intercept)", "Age", "Study Group", "Age:Study Group"),
  Estimate = c(4.1459, -0.1640, -1.6818, 0.1098),
  `Standard Error` = c(0.7175, 0.0751, 1.0279, 0.1024),
  `P-value` = c("< .0001", "0.0301", "0.1032", "0.2847"),
  check.names = FALSE
)

kable(regression_table, caption = "Linear Regression Model Output", format = "markdown")
```

Our findings indicate that age is a significant factor in FPS-R scores, as suggested by the p-value in Table 2. The model suggests that older children experience less pain during IV insertion. Specifically, with each additional year of age, the FPS-R scores decrease by an average of 0.1640 in the control group and by an average of 0.0542 in the study group.

Compared to the control group, children in the study group experienced a greater decrease in FPS-R scores by an average of 1.6818. This matches our earlier findings for the main objective that vapocoolant spray is effective in reducing pain levels in IV insertions for children.

Although the estimates for the randomization group and the interaction between age and the randomization group are not statistically significant, they still provide valuable insights. From the estimates in Table 2, a linear regression model is created which represents the average FPS-R scores among children according to their age and group allocation, as depicted in Figure 2.

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%", fig.align='center', fig.cap='Linear Regression Model for Control and Study Group'}
# Plot the linear regression line
ggplot(data, aes(x = Age, y = `FPS-R score`, color = factor(Randomization))) +
  # Add the regression line
  geom_smooth(method = "lm", se = FALSE) +
  # Set plot labels
  xlab("Age") +
  ylab("FPS-R score") +
  labs(color = "Randomization", title = "Linear Regression Model for Control and Study Group") +
  scale_color_manual(values = c("#219ebc", "#fb8500"),
                     labels = c("Control Group", "Study Group")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Figure 2 illustrates that, on average, the study group experiences less pain compared to the control group. However, the control group shows a much steeper decrease in FPS-R scores. As age increases, the gap in FPS-R score reductions between the study and control groups narrows, suggesting that the vapocoolant spray is more effective in reducing pain for younger children. Notably, the two lines intersect at around age 15, indicating that the efficacy of the vapocoolant spray in reducing pain may not be as pronounced in children aged 15 and 16.

\vspace{3mm}

**4.3 Secondary Objective 2: To explore the change in the number of IV insertion attempts when using Ametop alone versus the combination of Ametop with a vapocoolant spray**

Figure 3 below represents the number of IV insertion attempts in the control group compared to the study group with vapocoolant spray.

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%", fig.align='center', fig.cap = "Number of IV Attempts for Control and Study Group"}
# Bar Chart for IV attempts
bar <- data %>%
  mutate(Group = ifelse(Randomization == 1, "Control Group", "Study Group")) %>%
  ggplot(aes(x = `Number of IV attempts`, fill = Group)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Control Group" = "#219ebc", "Study Group" = "#fb8500")) +
  labs(title = "Number of IV Attempts for Control and Study Group", x = "Number of IV Attempts",
    y = "Count", fill = "Group"
  ) +
    theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

print(bar)
```

Figure 3 shows that, for the distribution of IV attempts, both groups exhibit comparable patterns, with the majority of instances in both groups having a single IV attempt. Quantitatively, the study group, with vapocoolant spray, has an average of 1.08 IV attempts. This is marginally lower than the control group’s average of 1.155 attempts, where only Ametop is applied. We thus conclude that there is no increase in the number of IV attempts for the study group when the vapocoolant spray is added.

In summary, through the analysis of simple summary plots and calculation of mean attempts, it becomes evident that the introduction of the vapocoolant spray does not result in an elevated number of IV attempts when combined with Ametop.

\vspace{3mm}

**4.4 Secondary Objective 3: To explore the incidence of side effects due to Ametop**

First, we want to know the proportions of patients experiencing side effects. Our findings indicate that there are 12 out of 240 patients from the dataset that experience side effects. This finding is shown in Figure 4 pie chart.

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%", fig.align='center', fig.cap = "Proportion of Observed Side Effects"}
# Pie chart draw function with percentages
draw_pie_chart <- function(all_categories, title) {
  # Count the occurrences of each category
  category_counts <- table(all_categories)

  # Calculate the percentage for each category
  percentages <- round(category_counts / sum(category_counts) * 100, 1)

  # Create labels that include both the category name and the percentage
  labels <- paste(names(category_counts), " - ", percentages, "%", sep="")

  pie(category_counts,
      main = title,
      col = c("#06d6a0", "#ef476f"),
      labels = labels,
      cex = 1.0)
}

# Summarize "Any side effects observed?" column
side_effects_observed_summary <- data.frame(original_data$`Any side effects observed?`)

# Create a pie chart for the observed side effects
draw_pie_chart(side_effects_observed_summary, "Observed Side Effects")
```

As illustrated in Figure 4, 95% of the patients experience no side effects, and 5% of patients experience side effects.

Then, we want to understand the type of side effects present. Table 3 below shows the categorization of the side effects based on the description given in the dataset.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
side_effects_data <- data.frame(
  `Side Effects Description` = c(
    "Slight redness of skin from ametop",
    "Some redness on skin, more than usual",
    "vey mild pink and pruritic skin.",
    "complained of pain on the spray.",
    "slightly puffy hands after ametop removal",
    "pink skin and very mild itching",
    "redness from ametop",
    "quite red & itchy skin from ametop. Underlying eczema.",
    "redness at ametop site. Not painful or pruritic.",
    "redness & raised area under ametop",
    "quite itchy from ametop. Redness.",
    "Slight redness from ametop. No pruritis."
  ),
  Category = c(
    "Redness",
    "Redness",
    "Redness + Itching",
    "Pain",
    "Puffiness",
    "Redness + Itching",
    "Redness",
    "Redness + Itching",
    "Redness",
    "Redness",
    "Redness + Itching",
    "Redness"
  ),
  check.names = FALSE
)

kable(side_effects_data, caption = "Side Effects Category")
```

The totals for each category in Table 3 are summarized in Table 4.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
side_effects_data <- data.frame(
  Category = c("Pain", "Puffiness", "Redness only", "Redness + Itching"),
  Frequency = c("1", "1", "6", "4")
)

kable(side_effects_data, caption = "Frequency Table of Side Effects")
```

As shown in Table 3 and Table 4, the most common side effect reported by patients is redness, either alone or accompanied by itching.

\newpage

# 5. Conclusion

**5.1 Conclusion**

The study investigates whether the addition of a vapocoolant spray significantly reduces IV insertion pain for pediatric patients compared to using only Ametop gel. The result shows that the combination of Ametop and vapocoolant spray significantly reduces discomfort in pediatric patients compared to Ametop alone. This combination is particularly effective in younger children and may not be necessary for children aged 15 and 16. Additionally, the addition of spray does not increase the number of IV attempts. In terms of side effects, 95% of patients report no adverse effects from Ametop. Among the 5% of patients experiencing side effects, the most common side effect is redness, either alone or accompanied by itching.

**5.2 Limitation**

The baseline pain level was not recorded during data collection. External factors, such as existing headaches or body aches, might have influenced the patients' perception of pain during IV insertion, potentially biasing their FPS-R scores.

Additionally, it is possible that several factors, such as age, the number of IV attempts, and the practitioner’s role, could significantly impact the FPS-R scores. However, restricting these variables within the current dataset would yield an insufficient amount of data for robust analysis.

**5.3 Recommendation**

Based on the limitation, future research should have a more controlled set of variables to enhance the understanding of how the vapocoolant spray affects IV insertion pain for pediatric patients. Studies could focus on children within a specific age group or a narrower age range, limiting other factors to a single IV attempt and practitioners with similar levels of experience. This approach would enable a more precise analysis and lead to more definitive conclusions about the efficacy of vapocoolant spray.

\newpage

# References
Rogers TL, Ostrow CL. The use of EMLA cream to decrease venipuncture pain in children. Journal of Pediatric Nursing. 2004 Feb 1;19(1):33-9.

\newpage

# Appendices

## For Client

### Two-Sample T-Test

Let $x_i$ be the value for one of the response variables in our sample. $x_1, x_2, ..., x_{223}$ are considered as IID sample of size n = 223.

> $H_0$: The mean FPS-R scores for the control and study groups are the same.

> $H_a$: The mean FPS-R score is lower for the study group which is using the vapocoolant spray.

$\bar{x}$ is the sample mean and is the estimate of the response variable, and the $s_x$ is the sample standard deviation.

The formula for the two-sample t-test is:

$$
t = \frac{\bar{X}_1 - \bar{X}_2}{\frac{s_x} {\sqrt{n}}}
$$

We calculated the result using a one-sided test in R, and it is displayed in Table 1.

### Linear Regression Model

Our model to test the relationship between age and randomization (control and study) groups on FPS-R scores is as follows:

$$
Y = 4.1459 - 0.1640X_1 - 1.6818X_2 + 0.1098X_1*X_2 + \epsilon
$$

where $\epsilon$ follows $N(0, \sigma^2)$, Y = FPS-R scores, $X_1$ = Age, and

$$
X_2 = \left\{ \begin{array}{cc} 1 & \text{if the patient is in the study group} \\ 0 & \text{if the patient is in the control group} \end{array}\right.
$$

From the linear regression, we can get the estimated coefficient and the p-value through R-codes that are presented in Table 2. Using the equation, we can calculate fitted values, which are the predicted values of the FPS-R scores using the data that is used to create the model, to create Figure 2.

\newpage

## For Mentor

### Data Cleaning

```{r, echo=TRUE}
# Read data
file_path <- "../data/Ametop-PE data(new-Alex)(1control 2study).xlsx"
original_data <- read_excel(file_path)

# Save original data for side effects analysis
data <- original_data

# Remove withdrawn patients
data <- data[data$`Withdrawn after consent?` != "Yes", ]

# Remove unnecessary variables
data <- select(data, 
               -`Elective surgery requiring IV?`,
               -`ASA I or II?`,
               -`Aged 5-16?`,
               -`Any allergies to Ametop, Pain Ease, or Tagederm adhesive?`,
               -`Ametop placed at least 30min before estimated IV start time?`,
               -`Receiving sedative pre-medication / anxiolytics?`,
               -`Needle phobia?`,
               -`Planned inhalation induction?`,
               -`Developmental delay or unable to interpret FPS-R?`,
               -`Time Ametop applied`,
               -`Time Ametop removed`,
               -`Time of skin puncture`,
               -`Record ID`,
               -`Complete?`,
               -`Withdrawn after consent?`,
               -`Reason for withdrawal`,
               -`Other reason for withdrawal`)
```

```{r, echo=TRUE}
# Check for missing values in all columns except 'Notes' and 'Observed side effects'
missing_values_summary <- data %>%
  select(-Notes, -`Observed side effects`) %>%
  summarise_all(~sum(is.na(.)))

# No missing values found
print(missing_values_summary)
```

```{r, echo=TRUE}
# Shorten the data
data$`Reaction to skin puncture` <- gsub("^(Slight|Severe).*", "\\1 Pain",
                                         data$`Reaction to skin puncture`)

# Verify the changes
table(data$`Reaction to skin puncture`)
```

### Data Preprocessing

```{r, echo=TRUE}
# Create new variables for the dataset

# Create 'Observer Number'
data$`Observer Number` <- case_when(
  data$`Reaction to skin puncture` == "None" ~ 1,
  data$`Reaction to skin puncture` == "Slight Pain" ~ 2,
  data$`Reaction to skin puncture` == "Severe Pain" ~ 3
)

# Create 'FPS-R Number'
data$`FPS-R Number` <- case_when(
  data$`FPS-R score` %in% c(0, 2) ~ 1,
  data$`FPS-R score` %in% c(4, 6) ~ 2,
  data$`FPS-R score` %in% c(8, 10) ~ 3
)

# Create 'Match Responses'
data$`Match Responses` <- ifelse(data$`Observer Number` == data$`FPS-R Number`, "Yes", "No")

# Create data subset for match data
data_subset <- data[data$`Match Responses` == 'Yes',]
```

```{r, echo=TRUE}
# Verify data
head(data)
```

### Main Objective

#### Boxplots

```{r, echo=TRUE, fig.align='center'}
# Boxplots
plot1 <- data %>% filter(Randomization == 1) %>%
  ggplot(aes(`FPS-R score`, fill = "Control Group")) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Control Group FPS-R Scores Boxplot") +
  scale_fill_manual(values = c("Control Group" = "#219ebc")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "none")

plot2 <- data %>% filter(Randomization == 2) %>%
  ggplot(aes(`FPS-R score`, fill = "Study Group")) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Study Group FPS-R Scores Boxplot") +
  scale_fill_manual(values = c("Study Group" = "#fb8500")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "none")

plot1
plot2
```

#### Table

```{r, echo=TRUE}
# Create a table to count matches
table <- table(data$`Observer Number`, data$`FPS-R Number`)

# Name the rows and columns to reflect the desired format
rownames(table) <- c("1", "2", "3")
colnames(table) <- c("1", "2", "3")

# Output the 3x3 table
cat("Contingency Table of FPS-R Number vs. Observer Number\n")
cat("Rows: FPS-R Number\n")
cat("Columns: Observer Number\n\n")
print(table)
```

#### T-tests

```{r, echo=TRUE}
# Perform t-test
t_test <- t.test(data[data$Randomization == 2, ]$`FPS-R score`,
                      data[data$Randomization == 1, ]$`FPS-R score`,
                      alternative = c("less"))
print(t_test)

# Conclusion: We should only reject the null hypothesis based on the full dataset.
```

### Secondary Objective 1: Linear Regression of Age * Randomization Group

```{r, echo=TRUE, fig.align='center'}
lm_age <- lm(`FPS-R score` ~ Age * factor(Randomization), data = data)
tidy(summary(lm_age))
plot(lm_age)
```

```{r, echo=TRUE, fig.align='center'}
# Plot the linear regression line
ggplot(data, aes(x = Age, y = `FPS-R score`, color = factor(Randomization))) +
  # Add the regression line
  geom_smooth(method = "lm", se = FALSE) +
  # Set plot labels
  xlab("Age") +
  ylab("FPS-R score") +
  labs(color = "Randomization", title = "Linear Regression Model for Control and Study Group") +
  scale_color_manual(values = c("#219ebc", "#fb8500"),
                     labels = c("Control Group", "Study Group")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

### Secondary Objective 2: Number of IV Attempts Analysis

```{r, echo=TRUE, fig.align='center'}
# Summary statistics
data_control <- data[data$`Randomization` == 1,]
data_study <- data[data$`Randomization` == 2,]
summary(data_control$`Number of IV attempts`)
summary(data_study$`Number of IV attempts`)

# Mean for study group decreases instead

# Bar Chart for IV attempts
bar <- data %>%
  mutate(Group = ifelse(Randomization == 1, "Control Group", "Study Group")) %>%
  ggplot(aes(x = `Number of IV attempts`, fill = Group)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Control Group" = "#219ebc", "Study Group" = "#fb8500")) +
  labs(title = "Number of IV Attempts for Control and Study Group", x = "Number of IV Attempts",
    y = "Count", fill = "Group"
  ) +
    theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

print(bar)

# No need further testing, from EDA we can see that the study group 
# didn't increase number of IV attempts (in fact, it decreases)
# (note: we did try doing t-test for confirmation and we can't reject null)
```

### Secondary Objective 3: Side Effects

#### Pie chart for Observed side effects

```{r, echo=TRUE, fig.align='center'}
# Pie chart draw function with percentages
draw_pie_chart <- function(all_categories, title) {
  # Count the occurrences of each category
  category_counts <- table(all_categories)

  # Calculate the percentage for each category
  percentages <- round(category_counts / sum(category_counts) * 100, 1)

  # Create labels that include both the category name and the percentage
  labels <- paste(names(category_counts), " - ", percentages, "%", sep="")
  
  pie(category_counts,
      main = title,
      col = c("#06d6a0", "#ef476f"),
      labels = labels,
      cex = 1.0)
}

# Summarize "Any side effects observed?" column
side_effects_observed_summary <- data.frame(original_data$`Any side effects observed?`)

# Create a pie chart for the observed side effects
draw_pie_chart(side_effects_observed_summary, "Observed Side Effects")
```

```{r, echo=TRUE}
# Define the categorization function
categorize_effects <- function(effect) {
  effect_lower <- tolower(effect)
  categories <- list(
    Redness = "red|redness|pink",
    Itching = "itchy|pruritic|pruritis|itching",
    Pain = "pain|painful",
    Puffiness = "puffy"
  )

  identified_categories <- character(0)

  # Negation patterns
  pattern_start <- "\\b(no|not)\\s+("
  pattern_middle <- ")\\b|(no|not)\\s+\\w+\\s+or\\s+("
  pattern_end <- ")"

  for (category_name in names(categories)) {
    symptom_pattern <- categories[[category_name]]

    # Check for the symptom but exclude if negated
    if (grepl(symptom_pattern, effect_lower)) {
      negation_pattern <- paste0(
        pattern_start, symptom_pattern, pattern_middle, symptom_pattern, pattern_end)
      if (!grepl(negation_pattern, effect_lower)) {
        identified_categories <- c(identified_categories, category_name)
      }
    }
  }

  # Return "Other" if no category is found
  if (length(identified_categories) == 0) {
    return("Other")
  }

  return(identified_categories)
}
```

#### Frequency table for observed side effects

```{r, echo=TRUE, fig.align='center'}
# Get the side effects data
side_effects_data <- na.omit(data.frame(SideEffects = original_data$`Observed side effects`))
side_effects_data$Category <- sapply(side_effects_data$SideEffects, function(effect) {
  category <- categorize_effects(effect)
  paste(category, collapse = " + ")
})
print(side_effects_data)

# Frequency table
table(side_effects_data$Category)
```